---
title: "Biostat 203B Homework 4 (Draft)"
subtitle: Due Mar 19 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
library(dplyr)
library(tidyverse)
library(lubridate)
library(miceRanger)
library(MatrixModels)
library(data.table)
library(rstatix)
library(glmnet)
library(randomForest)
library(caret)
library(Matrix)
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.

**Solution**:
MCAR: this is missing completely at random. The missing data values do not relate to any other data in the dataset and there is no pattern to the missing data values.
MAR: Missing at random - Missing data do have a relationship with the other variables in the dataset.The missing values themselves are random though.
MNAR: missing not at random, not only are the pattern of missing values is related to others in the dataset, the missing values are also not random as well.

2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.
**Solution**:
This replaces missing values by making certain assumptions about the missing values.  Makes the assumptions based on MCAR, MAR etc. until it converges and it will create however many datasets you choose. Then you can analyize however is needed. 

3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.
```{r}
icu_coh <-readRDS(str_c("/home/asburysean/biostat-203b-2021-winter/"
                        ,"Hw3/mimiciv_shiny/icu_cohort.rds"))
na_count <-sapply(icu_coh, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count
# Normally wouldnt drop all these but need space in order for it to run
drops <- c("first_careunit","last_careunit", "intime", "outtime", "los", 
           "admittime", "dischtime", "deathtime", "admission_type", 
           "admission_location", "discharge_location", "insurance", "language",
           "days_death", "arterial_blood_pressure_mean",
           "arterial_blood_pressure_systolic", "lactate", "dod", 
           "anchor_year_group", "anchor_year", "anchor_age", 
           "hospital_expire_flag", "edouttime", "edregtime", "hadm_id", 
           "stay_id")
icu_coh_tble <- icu_coh[ , !(names(icu_coh) %in% drops)]
print(icu_coh_tble)
colnames(icu_coh_tble)

icu_coh_tble$bicarbonate <- replace(icu_coh_tble$bicarbonate, 
                            is_extreme(icu_coh_tble$bicarbonate) == TRUE, NA)
icu_coh_tble$calcium <- replace(icu_coh_tble$calcium, 
                            is_extreme(icu_coh_tble$calcium) == TRUE, NA)
icu_coh_tble$chloride <- replace(icu_coh_tble$chloride, 
                            is_extreme(icu_coh_tble$chloride) == TRUE, NA)
icu_coh_tble$creatinine <- replace(icu_coh_tble$creatinine, 
                            is_extreme(icu_coh_tble$creatinine) == TRUE, NA)
icu_coh_tble$glucose <- replace(icu_coh_tble$glucose, 
                            is_extreme(icu_coh_tble$glucose) == TRUE, NA)
icu_coh_tble$magnesium <- replace(icu_coh_tble$magnesium, 
                            is_extreme(icu_coh_tble$magnesium) == TRUE, NA)
icu_coh_tble$potassium <- replace(icu_coh_tble$potassium, 
                            is_extreme(icu_coh_tble$potassium) == TRUE, NA)
icu_coh_tble$sodium <- replace(icu_coh_tble$sodium, 
                            is_extreme(icu_coh_tble$sodium) == TRUE, NA)
icu_coh_tble$hematocrit <- replace(icu_coh_tble$hematocrit, 
                            is_extreme(icu_coh_tble$hematocrit) == TRUE, NA)
icu_coh_tble$wbc <- replace(icu_coh_tble$wbc, 
                            is_extreme(icu_coh_tble$wbc) == TRUE, NA)
icu_coh_tble$heart_rate <- replace(icu_coh_tble$heart_rate, 
                            is_extreme(icu_coh_tble$heart_rate) == TRUE, NA)
icu_coh_tble$non_invasive_blood_pressure_systolic <- 
  replace(icu_coh_tble$non_invasive_blood_pressure_systolic, 
                            is_extreme(
                              icu_coh_tble$non_invasive_blood_pressure_systolic)
                           == TRUE, NA)
icu_coh_tble$non_invasive_blood_pressure_mean <- replace(icu_coh_tble$non_invasive_blood_pressure_mean, 
                            is_extreme(
                              icu_coh_tble$non_invasive_blood_pressure_mean)
                            == TRUE, NA)
icu_coh_tble$respiratory_rate <- replace(icu_coh_tble$respiratory_rate, 
                            is_extreme(icu_coh_tble$respiratory_rate)
                            == TRUE, NA)
icu_coh_tble$temperature_fahrenheit <- replace(
  icu_coh_tble$temperature_fahrenheit, 
  is_extreme(icu_coh_tble$temperature_fahrenheit) == TRUE, NA)
#Checking for unusual values
table(icu_coh_tble$marital_status)
table(icu_coh_tble$ethnicity)
table(icu_coh_tble$gender)
table(icu_coh_tble$age_adm)
table(icu_coh_tble$died_30days)



```



4. Impute missing values by `miceRanger` (request $m=3$ datasets). This step is very computational intensive. Make sure to save the imputation results as a file.
```{r}
# I had to use max.depth=10 due to computing power and it couldnt converge
seqTime <- system.time(
  miceObj <- miceRanger(
    icu_coh_tble
    , m=3
    , max.depth=10
    , returnModels = TRUE
    , verbose=FALSE
  )
)
miceObj
icucoh_new <- completeData(miceObj)
icucoh_new

```



5. Make imputation diagnostic plots and explain what they mean.
```{r}
# The red line is the density of the original nonmissing data and the black line is the density of the imputed data sets. You can see that there are some differences but and that could tell us if its MCAR. We had to use max.depth=10 so that could cause some issues.
plotDistributions(miceObj,vars='allNumeric')
# boxplot of the correlations between imputed values for each iteration and every combo. As you can see we have convergence issues due to computational power.
plotCorrelations(miceObj,vars='allNumeric')
# This shows if we have convergence issues and as you can see that we do have some due to computational power issues but noramally we would do more imputations if needed.
plotVarConvergence(miceObj,vars='allNumeric')
```


6. Obtain a complete data set by averaging the 3 imputed data sets.
```{r}
df1 <- icucoh_new$Dataset_1
df2 <- icucoh_new$Dataset_2
df3 <- icucoh_new$Dataset_3

#This works on my local but for some reason wasnt working on the server
# dfa <- data.frame(data.matrix(df1))
# dfb <- as.data.frame(data.matrix(df2, rownames.force = NA))
# dfc <- as.data.frame(data.matrix(df3, rownames.force = NA))
df1$ethnicity <- as.numeric(factor(df1$ethnicity))
df1$marital_status <- as.numeric((factor(df1$marital_status)))
df1$gender <- as.numeric(factor(df1$gender))
df2$ethnicity <- as.numeric(factor(df2$ethnicity))
df2$marital_status <- as.numeric((factor(df2$marital_status)))
df2$gender <- as.numeric(factor(df2$gender))
df3$ethnicity <- as.numeric(factor(df3$ethnicity))
df3$marital_status <- as.numeric((factor(df3$marital_status)))
df3$gender <- as.numeric(factor(df3$gender))
head(df3)
#Avg and bind, couldnt factorize marriage variable after avg since there were some decimals for some weird reason so kept it as continuous.
avgicu_coh <- rbindlist(list(df1, df2,df3))[,lapply(.SD,mean), list(subject_id)]
head(avgicu_coh)
avgicu_coh$ethnicity <- as.factor(avgicu_coh$ethnicity)
avgicu_coh$gender <- as.factor(avgicu_coh$gender)
avgicu_coh$died_30days <- as.factor(avgicu_coh$died_30days)
```



## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function), (2) logistic regression with lasso penalty (glmnet package), (3) random forest (randomForest package), or (4) neural network.

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.
```{r}

trainIndex <- createDataPartition(avgicu_coh$died_30days, p = .8, list = FALSE, times = 1)
head(trainIndex)
icuTrain <- avgicu_coh[ trainIndex,]
icuTest  <- avgicu_coh[-trainIndex,]



```


2. Train the models using the training set.
```{r}
#RandomForest
rf <- randomForest(
   died_30days~ .,
  data=icuTrain
)
rf
#logistic regression
# Dumy code categorical predictor variables
x <- model.matrix(died_30days~., icuTrain)[,-1]
# Convert the outcome (class) to a numerical variable
y <- icuTrain$died_30days
set.seed(123)
cv.lasso <- cv.glmnet(x, y, alpha = 1, family = "binomial")
plot(cv.lasso)
cv.lasso$lambda.min
cv.lasso$lambda.1se
coef(cv.lasso, cv.lasso$lambda.min)
lasso.model <- glmnet(x, y, alpha = 1, family = "binomial",
                      lambda = cv.lasso$lambda.min)
# Make prediction on test data
x.test <- model.matrix(died_30days ~., icuTest)[,-1]
probabilities <- lasso.model %>% predict(newx = x.test)
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
table(observed.classes)
# Model accuracy
observed.classes <- icuTest$died_30days
mean(predicted.classes == observed.classes)




```

3. Compare model prediction performance on the test set.
```{r}
#Model accuracy RandomForest
pred = predict(rf, newdata=icuTest[])
confusionMatrix(pred, icuTest$died_30days)

# Model accuracy logistic regression
observed.classes <- icuTest$died_30days
mean(predicted.classes == observed.classes)

```

